---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("cbsodataR")
```


```{r packages}
library(tidyverse)
library(readr)
library(rmarkdown)
library(yaml)
library(cbsodataR)
```

# Title Page

Diana
Ernest
Sofiia
Victoria
Yurii

Include the tutorial group number 

Include your tutorial lecturer's name

Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem
Education and Income Inequality in the Netherlands
Income inequality is an important societal concern in the Netherlands. Diferences in income influence various aspects of life, such as level of living, healthcare, education, therefore research demonstrated how individual's place of residence and educational background can predetermine socioeconomic prospects for the future. As a result, people with similar abilities may experience different economic opportunities depending on their educational background and geographic location.

Key factor associated with income is level of education. Higher education may lead to higher prospects for employment and bwtter wages. Regional exposure may also influence income through variations in labor market conditions, economic development, and access to educational resources. If income depends strongly on education level and region, policymakers can develop targeted measures to reduce inequalities and promote equal opportunities across different parts of the country.

In this report we quantify the relationship between level of education, region and income in the Netherlands using data from Statistics Netherlands (CBS). Specifically, we analyze how income varies across different education levels and regions  and assess whether these factors are associated with economic outcomes.

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}

#onderwijs <- cbs_get_data("85525NED")
#inkomen <- cbs_get_data("86162NED")

#write_csv(onderwijs, "data/onderwijs.csv")
#write_csv(inkomen, "data/inkomen.csv")

data.inkomen <- read.csv("data/inkomen.csv")
data.onderwijs <- read.csv("data/onderwijs.csv")

```


## 2.2 Short summary of the dataset(s)
```{r}
head(data.onderwijs)
```

### Dataset 1: Educational Attainment (85525NED)

The dataset "Bevolking; hoogstbehaald onderwijsniveau en regio" contains information on the highest educational attainment of people aged 15–75 in the Netherlands, broken down by region, gender, age group, and year. The data are collected by Statistics Netherlands (CBS) and are available from 2013–2024. The unit of observation is a combination of region and year. 

Key variables include: 
- the percentage of people with low education (primary education, VMBO, MBO1), - medium education (HAVO, VWO, MBO2-4), and higher education (HBO, WO). 

For this project, the share of highly educated people is the most relevant variable, as it can be compared with regional income levels. A limitation of the dataset is that it contains aggregated regional statistics rather than individual-level data, so relationships observed across regions may not necessarily apply to individuals.

```{r}
head(data.inkomen)
```

### Dataset 2: Personal Income (86162NED)

The dataset "Inkomen van personen; persoonskenmerken, regio" (CBS) contains information on personal income by region and demographic characteristics in the Netherlands. The unit of observation is a combination of region, year, gender, and personal characteristics.

Key variables include:
- RegioS: region code
- Perioden: year
- GemiddeldPersoonlijkInkomen: average personal income (1,000 euros)
- MediaanPersoonlijkInkomen: median personal income (1,000 euros)

The dataset covers the period 2011–2024. Figures for 2024 are provisional. A limitation of the dataset is that it contains aggregated regional data rather than individual-level observations.
For our analysis, the most relevant variable is median personal income or average personal income, because these can be compared with regional educational attainment. A limitation is that the dataset is aggregated by region, so it does not allow conclusions about individual people. Also, the 2024 figures are provisional, and regional averages may hide differences within municipalities or provinces.

###
These datasets can be merged using the region (RegioS) and year (Perioden) variables, allowing us to examine the relationship between educational attainment and personal income across Dutch regions.

| Dataset | Unit of observation | Key variables | Source |
|----------|----------|----------|----------|
| Onderwijs (85525NED) | One Dutch region in one year | Region, year, population size, share low education, share medium education, share high education | Statistics Netherlands (CBS StatLine), 2013–2024 |
| Inkomen (86162NED) | One Dutch region in one year | Region, year, average personal income, median personal income, number of persons with income | Statistics Netherlands (CBS StatLine), 2011–2024 |

``` r
inline_code = TRUE
```

## 2.3 Describe the type of variables included

Educational Attainment Dataset (data.onderwijs) (85525NED)

This dataset primarily contains socio-economic status (SES) information, as educational attainment is a common indicator of an individual's socio-economic position. The dataset includes variables on the highest completed level of education, population counts, age groups, gender, region, and year.

The data are obtained from administrative sources and registers maintained by Statistics Netherlands (CBS). Educational attainment is compiled using information from educational institutions and government records. The dataset is aggregated at the regional level, meaning that it reports percentages and population counts for groups rather than information on individual persons.

Income of Persons Dataset (data.inkomen) (86162NED)

The variables in this dataset mainly contain socio-economic status (SES) information, as they measure personal income levels across different regions and population groups. Key variables include average personal income, median personal income, the number of persons with income, region, year, gender, and other demographic characteristics.

The data are derived from administrative sources, such as tax records and social security registers, and are compiled by Statistics Netherlands (CBS). The information is therefore not collected through surveys or interviews. The dataset is aggregated by region and demographic group, meaning that it provides summary statistics rather than individual-level income data.

# Part 3 - Quantifying

## 3.1 Data cleaning

```{r missing values}
colSums(is.na(data.inkomen))
colSums(is.na(data.onderwijs))
#check for missing values
```
```{r dublictaes}
sum(duplicated(data.inkomen))
sum(duplicated(data.onderwijs))
#check for dublicates
```
```{r structure}
str(data.inkomen)
str(data.onderwijs)
```
```{r remove_na}
data.inkomen <- na.omit(data.inkomen)
data.onderwijs <- na.omit(data.onderwijs)
#remove no data
```

Filtering for data.inkomen

```{r income_cleaning}

data.inkomenclean <- data.inkomen %>%
  mutate(
    Geslacht = trimws(as.character(Geslacht)),
    Persoonskenmerken = trimws(as.character(Persoonskenmerken)),
    RegioS = trimws(as.character(RegioS)),
    Perioden = trimws(as.character(Perioden))
  ) %>%
  filter(
    Geslacht == "T001038",
    Persoonskenmerken == "53050",
    grepl("^LD", RegioS) |
      grepl("^PV", RegioS) |
      RegioS == "GM0363"
  ) %>%
  select(
    RegioS,
    Perioden,
    GemiddeldPersoonlijkInkomen_6,
    MediaanPersoonlijkInkomen_7
  )

head(data.inkomenclean)

```

Chosen variables:
RegioS - region identifier 
Perioden - time period 
GemiddeldPersoonlijkInkomen_6 - mean personal income (average)
MediaanPersoonlijkInkomen_7 - median personal income

Filtering for data.onderwijs
```{r filter_onderwijs}
data.onderwijsclean <- data.onderwijs %>%
  filter(
    trimws(Geslacht) == "T001038",
    Leeftijd %in% c(53050, 53700)
  ) %>%
  mutate(
    RegioS = trimws(RegioS),
    Perioden = trimws(Perioden)
  ) %>%
  group_by(RegioS, Perioden) %>%
  summarise(
    k_2HavoVwoMbo24_3 = mean(k_2HavoVwoMbo24_3, na.rm = TRUE),
    k_3HboWo_4 = mean(k_3HboWo_4, na.rm = TRUE),
    k_32HboWoMasterDoctor_9 = mean(k_32HboWoMasterDoctor_9, na.rm = TRUE),
    .groups = "drop"
  )
```
Chosen variables:
RegioS - region identifier 
Perioden - time period 
k_2HavoVwoMbo24_3 - Medium education variable
k_3HboWo_4 - Higher education variable
k_32HboWoMasterDoctor_9 - Bachelor + Master/Doctor variable
(Filtered with 53050 and 53700 age groups and took an average to not create identical tables)

Merging two data sets
```{r merged data sets}
data.combined <- data.inkomenclean %>%
  inner_join(data.onderwijsclean, by = c("RegioS", "Perioden"))

nrow(data.combined)
ncol(data.combined)
head(data.combined)
```


## 3.2 Generate necessary variables

1. variable: income gap
2. variable: high education share
```{r gen_var1}
data.combined <- data.combined %>%
  mutate(
    income_gap = GemiddeldPersoonlijkInkomen_6 -
                 MediaanPersoonlijkInkomen_7,

    high_education_share =
      k_3HboWo_4 +
      k_32HboWoMasterDoctor_9
  )
```


## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?




Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.



