---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
install.packages("tidyverse")
install.packages("yaml")
install.packages("rmarkdown")
install.packages("cbsodataR")
```


```{r packages}
library(tidyverse)
library(readr)
library(rmarkdown)
library(yaml)
library(cbsodataR)
```

# Title Page

Diana
Ernest
Sofiia
Victoria
Yurii

Include the tutorial group number

Include your tutorial lecturer's name

Simon Loop

# Part 1 - Identify a Social Problem

Use APA referencing throughout your document. [Here's a link to some explanation.](https://www.mendeley.com/guides/apa-citation-guide/)

## 1.1 Describe the Social Problem
Education and Income Inequality in the Netherlands
Income inequality is an important societal concern in the Netherlands. Diferences in income influence various aspects of life, such as level of living, healthcare, education, therefore research demonstrated how individual's place of residence and educational background can predetermine socioeconomic prospects for the future. As a result, people with similar abilities may experience different economic opportunities depending on their educational background and geographic location.

Key factor associated with income is level of education. Higher education may lead to higher prospects for employment and bwtter wages. Regional exposure may also influence income through variations in labor market conditions, economic development, and access to educational resources. If income depends strongly on education level and region, policymakers can develop targeted measures to reduce inequalities and promote equal opportunities across different parts of the country.

In this report we quantify the relationship between level of education, region and income in the Netherlands using data from Statistics Netherlands (CBS). Specifically, we analyze how income varies across different education levels and regions  and assess whether these factors are associated with economic outcomes.

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}

#onderwijs <- cbs_get_data("85525NED")
#inkomen <- cbs_get_data("84494NED")

#write_csv(onderwijs, "data/onderwijs.csv")
#write_csv(inkomen, "data/inkomen.csv")

data.inkomen <- read.csv("data/inkomen.csv")
data.onderwijs <- read.csv("data/onderwijs.csv")

```


## 2.2 Provide a short summary of the dataset(s)

```{r}
head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

``` r
inline_code = TRUE
```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

-   Do the variables contain health information or SES information?

-   Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}
mean(dataset$percollege)
```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r gen_var1}

```

Variable 2

```{r gen_var2}

```

## 3.3 Visualize temporal variation

```{r}

```

## 3.4 Visualize spatial variation

```{r visualise_map}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}
dataset$inmetro <- dataset$inmetro %>% as.factor()
# Boxplot of poverty rate by state using the 'midwest' dataset
ggplot(dataset, aes(x = inmetro, y = percadultpoverty)) +
  geom_boxplot() +
  labs(
    title = "Distribution of Poverty Rates by Metropolitan status (Midwest counties)",
    x = "Metropolitan Area",
    y = "Poverty Rate of Adults (%)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}

```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.
